Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-13 talk/talk.tex
Fiddle around with language.

\begin{frame}{coupling via operator}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item which is a regularity conjecture
    \end{itemize}
\end{itemize}

@@

    \begin{itemize}
    \item $\Phi$ is diffusive in the large
    \item $\Phi$ is degenerate at the ice margin
    \end{itemize}
\item yet common to say
    \begin{itemize}
    \item $\Phi$ is hyperbolic because SKE is a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{block}{numerical goal}
given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$, and convex set $\mathcal{K}^h = \{r^h \ge b^h\} \subset \mathcal{V}^h$, solve VI for $s^h \in \mathcal{K}^h$:

@@

\begin{columns}
\column{0.35\textwidth}
    \begin{itemize}
    \small
    \item which solution variable?
    \item surface elevation ($s$) or thickness ($H$)?
    \item interior smoothness means $s$ is better for multigrid
    \item bed roughness means \emph{not} having ``fake ice'' where ice-free ($s=b$)
    \end{itemize}
\column{0.65\textwidth}
    \includegraphics[width=\textwidth]{figs/giscross.png}
\end{columns}
\end{frame}

@@

\begin{itemize}
\item this strategy gives $O(m\log m)$ solver for Laplacian obstacle problem (Tai, 2003)
    \begin{itemize}
    \item using Gauss-Seidel smoother
    \item \only<1>{a multilevel VI strategy is the only hope for (near) optimality}\only<2>{{\color{ForestGreen} a multilevel VI strategy is the only hope for (near) optimality}}
    \item for nonlinear VI, need additional FAS strategy {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \end{itemize}
\item for IGP a single residual evaluation requires solving Stokes
    \begin{itemize}
    \item expensive but implemented {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \item Stokes is nonlocal

\smallskip
    \item Jacobian dense

\smallskip
    \item pointwise smoothers (GS, Jacobi) use diagonal entries
        \begin{itemize}
        \item<2>[{\color{black} $\circ$}] \alert{which apparently fail {\Large \, :-(}}
        \end{itemize}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Stokes on an extruded mesh}

@@

\begin{itemize}
\item for each base mesh level $\mathcal{T}^j$ in FAS V-cycle,
\item evaluate the residual $\Phi(s^j) - a^j$ in the VI:
    \begin{itemize}
    \item source $a^j$ arises from FAS formula
    \item extrude to current geometry $s^j$ (with or without cliffs)
    \item no extrusion where ice free\footnote{thanks to Lawrence Mitchell for enabling this Firedrake capability}
    \item solve Stokes problem $F_{\Lambda_{s^j}}(\bu^j,p^j)=0$ (Isaac et al, 2015)
    \item compute $\Phi(s^j) = - \bu^j|_{s^j} \cdot \bu_{s^j}$ and subtract $a^j$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \only<1>{now all the evil is in the smoother}\only<2>{\alert{now all the evil is in the smoother}}
\item options from weaker to stronger:
    \begin{itemize}
    \item nonlinear Richardson?
        \begin{itemize}
        \item[{\color{black} $\circ$}] not promising because degenerate
        \end{itemize}
    \item nonlinear pointwise smoothers (GS, Jacobi)?
        \begin{itemize}
        \item[{\color{black} $\circ$}] \only<1>{extremely expensive because residual nonlocal} \only<2>{\alert{extremely expensive because residual nonlocal}}
        \item[{\color{black} $\circ$}] seem to fail anyway
        \end{itemize}
    \item nonlinear \texttt{PCPatch} smoothers?
        \begin{itemize}
        \item[{\color{black} $\circ$}] what does this even mean?
        \end{itemize}
    \item exploit SIA analogy?
        \begin{itemize}
        \item[{\color{black} $\circ$}] pointwise smoothers work for SIA
        \end{itemize}
    \item reduced-space Newton using banded Jacobian approximation?
        \begin{itemize}
        \item[{\color{black} $\circ$}] finite-difference Jacobian entries using pseudo-coloring
        \end{itemize}
    \end{itemize}

\bigskip
\item<2> status report: \alert{nothing works well so far}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emph{GOAL 1}:\, no shallow approximations in flow physics {\Large \strut} \only<2-3>{{\Large \color{ForestGreen} $\checkmark$}}
    \begin{itemize}
    \item<2-3> my Stokes solver is working in Firedrake on extruded mesh
    \item<2-3> solver choices by (Isaac et al, 2015) putatively optimal
    \end{itemize}
\item \emph{GOAL 2}:\, beat explicit time-stepping and enable paleo {\Large \strut} \only<3>{\Large \alert{:-(}}
    \begin{itemize}
    \item<2-3> problem is
        \begin{itemize}
        \item<2-3>[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
        \item<2-3>[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
        \item<2-3>[{\color{black} $\circ$}] nonlocal (Stokes)
        \end{itemize}
    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for the VI}
    \item<3> \alert{stuck on finding a robust optimal smoother}
    \end{itemize}

>>>>

\begin{frame}{rewrite coupling as operator}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item which is a regularity conjecture for the coupled theory
    \end{itemize}
\end{itemize}

@@

    \begin{itemize}
    \item $\Phi$ is elliptic in the large
    \item $\Phi$ is degenerate at the ice margin
    \end{itemize}
\item yet common to say
    \begin{itemize}
    \item $\Phi$ is advective because SKE is write-able as a transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{block}{numerical problem}
given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$, and convex set $\mathcal{K}^h = \{r^h \ge b^h\} \subset \mathcal{V}^h$, solve VI for $s^h \in \mathcal{K}^h$:

@@

\begin{columns}
\column{0.35\textwidth}
    \begin{itemize}
    \small
    \item which solution variable?
        \begin{itemize}
        \item surface elevation $s$?
        \item thickness $H$?
        \end{itemize}
    \end{itemize}
\column{0.65\textwidth}
    \includegraphics[width=\textwidth]{figs/giscross.png}
\end{columns}

\bigskip
\begin{itemize}
\item \alert{smoothness suggests $s$ is better for multigrid}
    \begin{itemize}
    \item \emph{corollary}. do not put ``fake ice'' where $s=b$
    \item only solve for ice velocity where there is ice!
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item in V-cycle, on each level solve a VI:
    $$\ip{\Phi^j(s^j + z^j) - a^j}{v^j - z^j} \ge 0 \quad \text{ for all $v^j \ge \chi^j$}$$

\item this strategy gives an $O(m\log m)$ solver for the Laplacian obstacle problem (Tai, 2003)
    \begin{itemize}
    \item \only<1>{a multilevel VI strategy is the only hope for (near) optimality}\only<2>{\alert{a multilevel VI strategy is the only hope for (near) optimality}}
    \end{itemize}
\item for IGP:
    \begin{itemize}
    \item for nonlinear VI, need additional FAS strategy {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \item each residual is nonlocal and expensive {\Large\strut} \only<2>{{\color{ForestGreen} \, {\Large $\checkmark$}}}
    \item<2> {\color{ForestGreen} strategy implemented}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Stokes on a Firedrake extruded mesh}

@@

\begin{columns}
\column{0.77\textwidth}
\begin{itemize}
\item V-cycle based on mesh hierarchy $\{\mathcal{T}^j\}$ over $\Omega$
\item to evaluate the residual $\Phi(s^j) - a^j$ in the VI:
    \begin{itemize}
    \item extrude base mesh $\mathcal{T}^j$ to current geometry $s^j$
        \begin{itemize}
        \item[{\color{black} $\circ$}] with or without small cliffs
        \end{itemize}
    \item no extrusion where ice free
        \begin{itemize}
        \item[{\color{black} $\circ$}] thanks to Lawrence Mitchell for this capability
        \end{itemize}
    \item solve Stokes problem $F_{\Lambda_{s^j}}(\bu^j,p^j)=0$
    \item compute $\Phi(s^j) = - \bu^j|_{s^j} \cdot \bn_{s^j}$ and subtract $a^j$
    \end{itemize}
\end{itemize}
\column{0.23\textwidth}
\input{tikz/vcycle.tex}
\end{columns}
\end{frame}

@@

\begin{itemize}
\item now all the evil is in the smoother
\item smoother options from weaker to stronger:
    \begin{itemize}
    \item nonlinear Richardson?
        \begin{itemize}
        \item[{\color{black} $\circ$}] not promising because degenerate
        \end{itemize}
    \item nonlinear pointwise smoothers (GS, Jacobi)?
        \begin{itemize}
        \item[{\color{black} $\circ$}] \only<1>{GS extremely expensive because residual nonlocal} \only<2>{\alert{GS extremely expensive because residual nonlocal}}
        \item[{\color{black} $\circ$}] seem to fail
        \end{itemize}
    \item exploit SIA analogy?
        \begin{itemize}
        \item[{\color{black} $\circ$}] pointwise smoothers work for SIA
        \end{itemize}
    \item nonlinear and nonlocal \texttt{PCPatch} smoothers?
        \begin{itemize}
        \item[{\color{black} $\circ$}] what does this even mean?
        \end{itemize}
    \item reduced-space Newton using banded Jacobian approximation?
        \begin{itemize}
        \item[{\color{black} $\circ$}] finite-difference Jacobian entries using pseudo-coloring
        \end{itemize}
    \end{itemize}

\medskip
\item<2> status report:
    \begin{itemize}
    \item \alert{nothing works well so far}
    \item NewtonRS and SIA-based smoothers most promising?
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emph{GOAL 1}:\, no shallow approximations in flow physics {\Large \strut} \only<2-3>{{\Large \color{ForestGreen} $\checkmark$}}
    \begin{itemize}
    \item<2-3> my Stokes solver is working in Firedrake on extruded mesh
    \item<2-3> solver choices by (Isaac et al, 2015) putatively optimal
    \end{itemize}
\item \emph{GOAL 2}:\, beat explicit time-stepping and enable long-time paleo simulations {\Large \strut} \only<3>{\Large \alert{:-(}}
    \begin{itemize}
    \item<2-3> problem is
        \begin{itemize}
        \item<2-3>[{\color{black} $\circ$}] constrained (nonlinear VI on $\mathcal{K} = \{s\ge b\}$)
        \item<2-3>[{\color{black} $\circ$}] diffusive in the large (Stokes $\approx$ SIA)
        \item<2-3>[{\color{black} $\circ$}] nonlocal (Stokes)
        \end{itemize}
    \item<2-3> {\color{ForestGreen} MCD FAS path to optimality for solving the VI}
    \item<3> \alert{stuck on finding a robust optimal smoother}
    \end{itemize}
